Novel genetic control of migratory diapause in Australian monarch butterflies

### 2025-03-13 paper/paper_draft.Rmd
changes in prep for resubmission. Still needs cleanup

---
title: "Novel genetic control of migratory diapause in Australian Monarch Butterflies"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  word_document:
    reference_docx: "paper-files/style_ref.docx"
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    fig_width: 7
    fig_height: 6
bibliography: "paper-files/citations.bib"
csl: "paper-files/apa.csl"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespacing
  - \usepackage{sectsty}\sectionfont{\fontsize{12}{12}\selectfont}
  - \usepackage{sectsty}\subsectionfont{\normalfont\itshape\fontsize{12}{12}\selectfont}
  - \usepackage[round]{natbib}
indent: true
sansfont: Times New Roman
fontsize: 12pt
always_allow_html: true
---

@@

@@

@@

# generate plot
t.sizes <- c(18, 12)
Figure2a <- ggplot(bias_p, aes(x = rank, y = score, size = total)) +
  geom_point(color ="#393e42") + theme_bw() +
  geom_line(data = perm_summary, aes(x = rank, y = score), inherit.aes = FALSE, color = "#eb612e") +
  geom_ribbon(data = perm_summary, aes(ymin = lCI, ymax = uCI, x = rank), 
              inherit.aes = FALSE, color = "#eb612e", alpha = .2) +
  xlab("Maternal Family") +
  ylab("Mean Reproductive Score") +
  labs(size = "Family Size")  +
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1])) +
  scale_size_continuous(range = c(5, 10))

@@

q_quant <- plot_qq(adat_quant, "pval")
Figure2b <- ggplot(q_quant$.base$data, aes(x = .e, y = .o)) +
  geom_point(color = "#393e42") +
  geom_abline(slope = 1, intercept = 0, color = "#eb612e") +
  theme_bw() +
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1])) +
  ylab(bquote(Observed ~ log[10](p))) +
  xlab(bquote(Expected ~ log[10](p)))

@@

# plot
Figure2c <- Figure2c +
  ylab("-log10(p-value)") +
  xlab("Chromosome") +
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1])) +
  scale_x_continuous(expand = c(0, 0), breaks = chr.centers, labels = 
                       names(chr.centers))

@@

#==========combine=======================
Figure2_top <- plot_grid(Figure2a, Figure2b, align = "hv", axis = "tl")

@@

#========plot====================
plot_parts <- vector("list", nrow(targets))
for(i in 1:nrow(targets)){
  # figure out max and min range
  min_pos <- targets[i,start]
  max_pos <- targets[i,end]

@@

  pt <- ggplot(tsnps, aes(x = Position/1000000, y = -log10(fdr))) +
    geom_point() +
    theme_bw() +
    scale_x_continuous(limits = c(min_pos/1000000, max_pos/1000000)) +
    scale_y_continuous(expand = c(0, .05))

@@

  pb <- ggplot(tgtf, aes(xmin = start/1000000, xmax = end/1000000, 
                         y = OGS_name, forward = direction, fill = gene)) +
    geom_gene_arrow(arrow_body_height = unit(1, "lines"),
                    arrowhead_height = unit(1.5, "lines"), show.legend = FALSE,
                    fill = "#E5C16F") + 
    geom_subgene_arrow(data = subgene, 
                       aes(xmin = gene_start/1000000, xmax = gene_end/1000000, 
                           xsubmin = start/1000000, xsubmax = end/1000000,
                           y = OGS_name), 
                       fill = "black", arrow_body_height = unit(1, "lines"),
                       arrowhead_height = unit(1.5, "lines")) +
    theme_bw() +
    scale_x_continuous(limits = c(min_pos/1000000, max_pos/1000000)) +
    ylab("Gene") +
    xlab(paste0("Chromosome ", targets[i]$chr, "Position (kb)"))

@@

<!-- Introduction: -->
<!--   Migration, partial migration, and adaptation to environmental change. -->
<!--   Monarchs in the Pacific and Australia, History -->
<!--   Two options for migratory behavior in Aus (proximate vs ultimate causes as well) -->
<!--   what we did -->

@@

Monarch Butterflies (Danaus plexippus) are a charismatic and culturally important North American butterfly species that are famous for their dramatic and unique migratory life history. While non-migratory populations of the species are widespread and not of particular conservation concern, migratory populations in North America have recently seen dramatic declines, prompting concern that the migratory life history of the species may vanish over the coming years as selective factors begin to favor residency over migration. However, one relatively new non-native monarch population in Australia rapidly re-acquired a migratory life history following hundreds of generations of residency and successive bottlenecks as the species stepping-stoned across the Pacific during the late 1800's and early 1900's. The genetic process by which this happened is not currently known. We raised and sequenced individuals from Queensland, Australia and found a novel association between reproductive diapause, a key migratory trait, and a genetic locus that has not previously been identified to be involved in migration in North America. This may imply that the species might be able to rapidly re-evolve migration in other places as well and thus provides some hope that the feared loss of migration in North American monarchs may not be permanent.

@@

Since they rely on multiple distinct ecosystems and consistent seasonal timings, migratory species are acutely at risk from global anthropogenic environmental change [@Both2010; @chowdhurySeasonalSpatialDynamics2021].The loss of breeding, wintering, or any transitional habitats may all cause declines in migrants [@Robbins1989], as may shifts in resource phenology due to seasonal changes which can cause resource availability and species life-histories to fall out of synchrony [@Jones2010a]. It is therefore not surprising that widespread declines have been observed in many migratory taxa [@Limburg2009;@Robbins1989;@rungeConservingMobileSpecies2014].

Population-level adaptability in migratory life-history should therefore be expected to increase long-term population viability by conferring resistance to phenological and environmental changes [@Pulido7341]. Migratory life-history adaptation can essentially occur along three distinct axes: 1) *temporal*, or adaptation that changes the phenology of migration; 2) *spatial*, or adaptation that alters migratory pathways and orientation; and 3) *residency*, or adaptation that shifts populations between migratory and resident life histories. Inter-population variation and adaptation in each of these pathways is well known from nature: temporal changes or variation in migration are well documented in many taxa including as birds [@Rubolini2007; @Gordo2007; @chuCommunityScienceReveals2022; @prytulaRapidAdvancementSpring2022; @pinszkeLongtermChangesAutumn2023; @mondain-monvalFlywayscaleAnalysisReveals2021], fish [@kovachGeneticChangeEarlier2012], mammals [@shuertDecadalMigrationPhenology2022], and butterflies [@culbertsonLongtermMonitoringIndicates2022; @fitchettAdvanceTimingAnnual2022];  as is short-stopping, extension of migratory pathways, or complete re-direction of migration [@Ambrosini2011; @LaSorte2007; @Visser2009; @sonnleitnerRapidShiftsMigration2022;@rushingMigratoryBehaviorWinter2020;@dufourYellowbrowedWarblerPhylloscopus2022;@dufourNewWestwardMigration2021;@jiaShiftingMigrationRoute2021]; and shifts towards residency or partial migration have been observed and/or predicted in birds [@Pulido7341; @Plummer2015; @Møller2014;@haranPatternsPartialMigration2021; @falchiInterIntrapopulationVariability2023;@bonnetlebrunEffectsUrbanizationBird2020], butterflies [@Howard2010], mammals [@VanDerRee2006; @Fleming2003; @Jones2014a], fish [@adamsOpinionPieceEvolutionary2022], and other taxa.

Adaptation via a transition to residency is of both particular interest and concern for the conservation of migratory species. While switching to a resident life history can allow species to avoid the increasing risks of migration and the intrinsically large energetic costs of migration in general [@McWilliams2004; @Rankin1992; @Wikelski2003], there is reasonable concern that the "extinction" of migration may not be easily reversible in some species. This is concerning since migratory life-history variation is expected to increase population adaptability [@Pulido2007], and thus the permanent loss of migration could potentially compromise long-term population persistence. Beyond this, migratory species often provide critical ecosystem services [@lopez-hoffmanEcosystemServicesTransborder2017;@semmensAccountingEcosystemServices2011] and have cultural and/or ecological  significance [@burgerComplexityBioindicatorSelection2015;@schuetzCharacterizingCulturalNiches2019;@semmensQuantifyingEcosystemService2018;@reynoldsMigratorySpeciesEcological2011] that may not survive the transition to permanent residency.

The loss of migratory behavior is a major concern for monarch butterflies (*Danaus plexippus*) [@browerAnimalMigrationsEndangered1991;@satterfieldMigratoryMonarchsThat2018]. Specifically, North American migratory monarch populations have experienced severe declines over the last several decades, sparking a fear that those populations may go extinct in the near future [@Semmens2016]. However, many individual butterflies ostensibly from the North American migratory population are now year-round residents in some areas in the southern United States [@Howard2010], where they can survive due to the presence year-round of introduced tropical milkweed (*Asclepias curassavica*) [@bataldenPotentialChangesEastern2015], and there are wide-spread, non-migratory monarch populations that are not of any immediate conservation concern [@walkerDanausPlexippusErrata]. While the species as a whole may therefore not be at risk, their migratory behavior almost certainly is. In addition to the damage to the long-term population persistence that could result from the loss of migration in North American populations (as discussed above), the loss of the publicly beloved annual monarch migration and concomitant overwintering would be culturally costly [@Gustafsson2015].

Interestingly, however, there is a population of monarchs in Australia in which migratory behavior has apparently re-emerged after it was previously lost. This population, which was established sometime before 1871 [@Zalucki2004], is the product of a series of sequential introductions that crossed the Pacific Ocean from North America beginning sometime in the early 1840s after the introduction of their host plants enabled individuals to survive after being blown from the mainland [Figure 1, see @Zalucki2004; @Pierce2014; @Hemstrom2022a]. Monarchs are not known to be migratory on any Pacific islands, and no migratory behavior was observed in Australia until the 1930s, when non-reproductive overwintering aggregations were observed in northern New South Wales [NSW, see @James2019]. These and other individuals were later shown to migrate hundreds of kilometers each year to their wintering grounds from more southerly, inland locations in NSW [@James2019;@jamesInductionReproductiveDormancy1983]. After over 90 years (and hundreds of generations) without any apparent migration, the behavior re-emerged in these populations. Understanding how this happened could provide crucial context around the "extinction" of migration in monarchs and other species.

To better understand how migration re-emerged in Australian monarchs, we studied continuously breeding monarch populations in Queensland. While these populations do not experience substantial seasonal fluctuations in resource availability, and thus do not migrate [@James2019], recent work has shown that reproductive diapause can be induced in some, *but not all* individuals from these populations if they are exposed to a reduction in photoperiod during development [@Freedman2018], a trait that these non-migratory populations do not display in the wild. In brief, reproductive diapause, where individuals delay their reproductive maturation, is a critical and tightly linked phenotype for migratory monarchs because it allows them to invest more heavily in lipid storage and increase their average longevity [@Herman2001]. Like other migratory behaviors, diapause is initiated by changes in day length, temperature, and other seasonal fluctuations [@Goehring2002;@iiamsPhotoperiodicClockRegulation2019] and is thought to be directly controlled by juvenile hormone (JH) concentration during development [@GreenII2019;@Herman2001].

Diapause control is also an phenotype of interest because it highlights a theoretical dichotomy in the types of genes that are associated with migratory life histories: migratory genes can either be directly causal, proximate migratory genes that are required prior for migration to occur (such as genes that control diapause, directional motion, and navigation), or facultative migratory genes that increase fitness in already migratory populations but are not strictly required for migratory behavior to occur (such as genes that change wing or limb shapes, respiration, lipid storage, and muscle strength). For a newly migratory population, like monarchs in Queensland, we would expect to the strongest genetic associations with migration to occur in proximate migratory genes that cause traits like diapause rather than facultative genes which may take longer to increase in frequency following the evolution of migratory behavior.

The variance in diapause induction in Queensland monarchs allows for a direct search for genetic associations with this trait and thus for an examination of the genetic underpinnings of the re-evolution of migration in Australian monarchs. Here, we used reduced representation sequencing to search for genes associated with migratory diapause in Queensland monarch butterflies. We found one strongly associated locus co-located with an E3 ubiquitin ligase gene. This gene has been previously identified as associated with intermolt transitions in silk moths (Bombyx mori), which is controlled by the same major hormone (juvenile hormone, hereafter JH) as reproductive diapause. This gene had not previously been identified as associated with migration in North American monarch butterflies, and we did not find that any previously identified migratory genes were associated with diapause in our study.

# Materials and Methods
## Sample Collection, Incubation, and Phenotyping

@@

## Statistical Analysis

>>>>

---
title: "Novel genetic control of migratory diapause in Australian monarch butterflies"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  word_document:
    reference_docx: "paper-files/style_ref.docx"
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    fig_width: 7
    fig_height: 6
bibliography: "paper-files/citations.bib"
csl: "paper-files/vancouver.csl"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespacing
  - \usepackage{sectsty}\sectionfont{\fontsize{12}{12}\selectfont}
  - \usepackage{sectsty}\subsectionfont{\normalfont\itshape\fontsize{12}{12}\selectfont}
  - \usepackage[round]{natbib}
indent: true
sansfont: Times New Roman
fontsize: 12pt
always_allow_html: true
---

@@

library(rnaturalearth)
library(ggplot2)
library(sf)
library(ggspatial)
library(ggimage)

@@

```{r base_map, include=FALSE}
world <- ne_countries(scale = 'small', returnclass = 'sf')

polygon <- st_polygon(x = list(rbind(c(-0.0001, 90),
                                     c(0, 90),
                                     c(0, -90),
                                     c(-0.0001, -90),
                                     c(-0.0001, 90)))) |>
  st_sfc() |>
  st_set_crs(4326)

sf::sf_use_s2(FALSE)

world <- world |> st_difference(polygon)

world_robin <- st_transform(world, crs = '+proj=robin +lon_0=180 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs')

sf::sf_use_s2(TRUE)

# world <- ggplot() +
#   geom_sf(data = world_robin, fill= "azure2")  + #background color for land
#   coord_sf(xlim = c(-70e5, 100e5),
#            ylim = c(-60e5, 58e5)) +
#   theme_bw() +
#   annotation_scale(text_cex = 1.5) + 
#   theme(axis.text = element_text(size = 20),
#         axis.title = element_text(size = 24)) +
#   annotation_north_arrow(style = north_arrow_fancy_orienteering(text_size = 20), location = "bl", 
#                          pad_y = grid::unit(.05, "native"))
# 
# world
# ggsave("results/map.jpg", world, "jpg", width = 15, height = 11) # note, need to annotate this with ranges and arrows to produce final image

world <- ggplot() +
  geom_sf(data = world_robin, fill= "#dec079")  + #background color for land
  coord_sf(xlim = c(-70e5, 100e5),
           ylim = c(-60e5, 58e5)) +
  theme_bw() +
  annotation_scale(text_cex = 1.5, bar_cols = c("#eb612e", "#dec079"), text_family = "serif", text_col = "#393e42", pad_y = unit(0.75, "cm")) +
  theme(axis.text = element_text(size = 20, color = "#393e42"),
        axis.title = element_text(size = 24),
        panel.grid = element_line(color = "#393e42")) +
  annotation_north_arrow(style = north_arrow_fancy_orienteering(text_size = 20,
                                                                text_col = "#393e42",
                                                                text_family = "serif",
                                                                fill = c("#eb612e", "#dec079"),
                                                                line_col = "#393e42"),
                         location = "bl",
                         pad_y = grid::unit(.05, "native"))

world
ggsave("Figure1_base.jpg", world, "jpg", width = 15, height = 11)
```

@@

# generate plot
t.sizes <- c(18, 12)
Figure2a <- ggplot(bias_p, aes(x = rank, y = score, size = total)) +
  geom_point(color ="#393e42") + theme_bw() +
  geom_line(data = perm_summary, aes(x = rank, y = score), inherit.aes = FALSE, color = "#eb612e") +
  geom_ribbon(data = perm_summary, aes(ymin = lCI, ymax = uCI, x = rank), 
              inherit.aes = FALSE, color = "#eb612e", alpha = .2) +
  xlab("Maternal Family") +
  ylab("Mean Reproductive Score") +
  labs(size = "Family Size")  +
  theme(legend.text = element_text(size = t.sizes[2] - 2),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[2]),
        title = element_text(size = t.sizes[1] + 2)) +
  scale_size_continuous(range = c(5, 10)) +
  ggtitle("B")

@@

q_quant <- plot_qq(adat_quant, "pval")
Figure2b <- ggplot(q_quant$.base$data, aes(x = .e, y = .o)) +
  geom_point(color = "#393e42") +
  geom_abline(slope = 1, intercept = 0, color = "#eb612e") +
  theme_bw() +
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1]),
        title = element_text(size = t.sizes[1] + 2)) +
  ylab(bquote(Observed ~ log[10](p))) +
  xlab(bquote(Expected ~ log[10](p))) +
  ggtitle("C")

@@

# plot
Figure2c <- Figure2c +
  ylab("-log10(p-value)") +
  xlab("Chromosome") +
  theme(legend.text = element_text(size = t.sizes[2]),
        axis.text = element_text(size = t.sizes[2]),
        axis.title = element_text(size = t.sizes[1]),
        axis.text.y = element_text(size = t.sizes[2]),
        legend.title = element_text(size = t.sizes[1]),
        title = element_text(size = t.sizes[1] + 2)) +
  scale_x_continuous(expand = c(0, 0), breaks = chr.centers, labels = 
                       names(chr.centers)) +
  ggtitle("D")

@@

#==========egg photos (actual A)=========
imgs <- data.frame(category = c("Mature", "Partially Mature", "Immature"),
                   img = c("~/GitHub/aus_monarchs/paper/paper-files/mature.JPG",
                           "~/GitHub/aus_monarchs/paper/paper-files/partially_mature.JPG",
                           "~/GitHub/aus_monarchs/paper/paper-files/immature.JPG"),
                   x = 1,
                   y = 1)
imgs$category <- factor(imgs$category, levels = c("Mature", "Partially Mature", "Immature"))

eggplot <- ggplot(imgs, aes(x = x, y = y, image = img)) +
  geom_image(size = 1.3) +
  facet_wrap(~category, ncol = 1, strip.position = "left") +
  theme_minimal() +
  theme(strip.text = element_text(size = t.sizes[2], angle = 90),
        strip.placement = "outside",
        title = element_text(size = t.sizes[1] + 2),
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank()) +
  ggtitle("A")

#==========combine=======================
f2al <- ggpubr::get_legend(Figure2a)
Figure2_top <- plot_grid(eggplot, Figure2a + guides(size = "none"), 
                         f2al, Figure2b, 
                         align = "hv", axis = "tlbr",
                         nrow = 1,
                         rel_widths = c(.5, 1, .2, 1)) # note, A is actually egg pictures, these labels are off

@@

#========plot====================
plot_parts <- vector("list", nrow(targets))
pmax <- max(-log10(adat_quant$fdr))

for(i in 1:nrow(targets)){
  # figure out max and min range
  min_pos <- targets[i,start]
  max_pos <- targets[i,end]

@@

  pt <- ggplot(tsnps, aes(x = Position/1000000, y = -log10(fdr))) +
    geom_point() +
    theme_bw() +
    scale_x_continuous(limits = c(min_pos/1000000, max_pos/1000000)) +
    scale_y_continuous(expand = c(0, .05), limits = c(0, pmax))

  pmax <- max(c(pmax, max(-log10(tsnps$fdr))))

@@

  tgtf$OGS_name[tgtf$OGS_name == "DPOGS204613"] <- paste0("Karst/DPOGS204613")
  subgene$OGS_name[subgene$OGS_name == "DPOGS204613"] <- paste0("Karst/DPOGS204613")

  pb <- ggplot(tgtf, aes(xmin = start/1000000, xmax = end/1000000, 
                         y = OGS_name, forward = direction, fill = gene)) +
    geom_gene_arrow(arrow_body_height = unit(1, "lines"),
                    arrowhead_height = unit(1.5, "lines"), show.legend = FALSE,
                    fill = "#E5C16F") + 
    geom_subgene_arrow(data = subgene, 
                       aes(xmin = gene_start/1000000, xmax = gene_end/1000000, 
                           xsubmin = start/1000000, xsubmax = end/1000000,
                           y = OGS_name), 
                       fill = "black", arrow_body_height = unit(1, "lines"),
                       arrowhead_height = unit(1.5, "lines")) +
    theme_bw() +
    scale_x_continuous(limits = c(min_pos/1000000, max_pos/1000000)) +
    ylab("Gene") +
    xlab(paste0("Chromosome ", targets[i]$chr, " Position (mb)"))

@@

```{r mapping_stats}
sort_flagstat <- list.files("../data/v4/flagstat/", "\\.sort\\.flagstat", full.names = TRUE)
flt_flagstat <- list.files("../data/v4/flagstat/", "\\.sort\\.flt\\.flagstat", full.names = TRUE)

sort_flagstat <- unlist(lapply(sort_flagstat, readLines))
flt_flagstat <- unlist(lapply(flt_flagstat, readLines))

mapping <- data.table(sample = sample.meta$SampleID,
                      reads = as.numeric(gsub(" .+", "", sort_flagstat[seq(1, length(sort_flagstat), by = 16)])),
                      mapped = as.numeric(gsub(" .+", "", sort_flagstat[seq(7, length(sort_flagstat), by = 16)])),
                      retained = as.numeric(gsub(" .+", "", flt_flagstat[seq(1, length(sort_flagstat), by = 16)])))

total_reads <- sum(mapping$reads)
total_mapped <- sum(mapping$mapped)
total_retained <- sum(mapping$retained)

percent_mapped <- total_mapped/total_reads
percent_retained <- total_retained/total_reads

```{r Table_S1}

TS1 <- all_genes[,c("gene_id", "Monarch geneid", "seqid", "start", "end")]
TS1$Chromosome <- targets$chr[match(TS1$seqid, targets$Chromosome)]
colnames(TS1) <- c("RefSeq ID", "OGS ID", "Chromosome RefSeq ID", "Start", "End", "Chromosome")
TS1 <- TS1[,c("RefSeq ID", "OGS ID", "Chromosome", "Chromosome RefSeq ID", "Start", "End")]
TS1 <- dplyr::arrange(TS1, Chromosome, Start)
fwrite(TS1, "TableS1.txt", sep = "\t", col.names = TRUE, row.names = FALSE, quote = FALSE)

@@

Monarch butterflies (*Danaus plexippus*) are a charismatic and culturally important North American butterfly species famous for their dramatic and unique migratory life history. While non-migratory populations of the species are widespread and apparently stable, migratory populations in North America have recently seen declines, prompting concern that the migratory phenomenon in North America may be at risk of disappearing. In contrast, a relatively recently-established monarch population in Australia has rapidly re-acquired a migratory life history following hundreds of generations of residency and successive bottlenecks as the species island-hopped across the Pacific during the late 1800s and early 1900s. The process by which migration re-emerged in Australian monarchs is not currently known. We raised and sequenced individuals from Queensland, Australia under environmental conditions associated with migration initiation and found a novel association between reproductive diapause, a key migratory trait, and the spectrin beta chain protein *Karst*. This protein is known to be involved in reproductive development in monarchs but has not previously been identified to be associated with migratory life history variance. Our results provide an intriguing example of how the temporary loss of migration—in this case likely over hundreds of generations—may not entail the loss of genetic variation associated with this complex life history strategy.

@@

Migratory species are acutely at risk from global anthropogenic environmental change [@Both2010; @chowdhurySeasonalSpatialDynamics2021]. The loss of breeding, wintering, or any transitional habitats may all cause declines in migrants [@Robbins1989], as may shifts in resource phenology due to seasonal changes which can cause resource availability and species life-histories to fall out of synchrony [@Jones2010a]. It is therefore not surprising that widespread declines have been observed in many migratory taxa [@Limburg2009;@Robbins1989;@rungeConservingMobileSpecies2014].

Population-level adaptability in migratory life-history should therefore be expected to increase long-term population viability by conferring resistance to phenological and environmental changes [@Pulido7341]. Migratory life-history adaptation can occur along three distinct axes: 1) *temporal*, or adaptation that changes the phenology of migration; 2) *spatial*, or adaptation that alters migratory pathways and orientation; and 3) *residency*, or adaptation that shifts populations between migratory and resident life histories. Inter-population variation and adaptation in each of these pathways is well known from nature: temporal changes or variation in migration are well documented in many taxa including as birds [@Rubolini2007; @Gordo2007; @chuCommunityScienceReveals2022; @prytulaRapidAdvancementSpring2022; @pinszkeLongtermChangesAutumn2023; @mondain-monvalFlywayscaleAnalysisReveals2021], fish [@kovachGeneticChangeEarlier2012], mammals [@shuertDecadalMigrationPhenology2022], and butterflies [@culbertsonLongtermMonitoringIndicates2022; @fitchettAdvanceTimingAnnual2022];  as is short-stopping, extension of migratory pathways, or complete re-direction of migration [@Ambrosini2011; @LaSorte2007; @Visser2009; @sonnleitnerRapidShiftsMigration2022;@rushingMigratoryBehaviorWinter2020;@dufourYellowbrowedWarblerPhylloscopus2022;@dufourNewWestwardMigration2021;@jiaShiftingMigrationRoute2021]; and shifts towards residency or partial migration have been observed and/or predicted in birds [@Pulido7341; @Plummer2015; @Møller2014;@haranPatternsPartialMigration2021; @falchiInterIntrapopulationVariability2023;@bonnetlebrunEffectsUrbanizationBird2020], butterflies [@Howard2010], mammals [@VanDerRee2006; @Fleming2003; @Jones2014a], and fish [@adamsOpinionPieceEvolutionary2022].

Adaptation via a transition to residency is of both particular interest and concern for the conservation of migratory species. While switching to a resident life history can allow species to avoid the increasing risks of migration and the intrinsically large energetic costs of migration in general [@McWilliams2004; @Rankin1992; @Wikelski2003], there is reasonable concern that the "extinction" of migration may not be easily reversible in some species. Permanent loss of migratory behavior is concerning because, in general, variation in migratory tendency is expected to increase long-term population persistence [@Pulido2007]. Beyond this, migratory species often provide critical ecosystem services [@lopez-hoffmanEcosystemServicesTransborder2017;@semmensAccountingEcosystemServices2011] and have cultural and/or ecological significance [@burgerComplexityBioindicatorSelection2015;@schuetzCharacterizingCulturalNiches2019;@semmensQuantifyingEcosystemService2018;@reynoldsMigratorySpeciesEcological2011] that may not survive the transition to permanent residency.

The loss of migratory behavior is a major concern for monarch butterflies (*Danaus plexippus*) [@browerAnimalMigrationsEndangered1991;@satterfieldMigratoryMonarchsThat2018]. Specifically, North American migratory monarch populations have experienced pronounced declines in counts of overwintering individuals over the last several decades, sparking a fear that those populations may go extinct in the near future [@Semmens2016; but see @crossleyOpposingGlobalChange2022]. However, many individual butterflies ostensibly from the North American migratory population are now year-round residents in some areas in the southern United States [@Howard2010], where they can survive due to the presence year-round of introduced tropical milkweed (*Asclepias curassavica*) [@bataldenPotentialChangesEastern2015], and there are widespread, non-migratory monarch populations that are not of any immediate conservation concern [@walkerDanausPlexippusErrata]. While the species as a whole may therefore not be at risk, their migratory behavior seemingly is. In addition to the damage to the long-term population persistence that could result from the loss of migration in North American populations (as discussed above), the loss of the publicly beloved annual monarch migration and concomitant overwintering would be culturally costly [@Gustafsson2015].

In contrast, there is a population of monarchs in Australia in which migratory behavior has apparently re-emerged after it was previously lost. This population, which was established sometime before 1871 [@Zalucki2004], is the product of a series of sequential introductions that crossed the Pacific Ocean from North America beginning sometime in the early 1840s after the introduction of their host plants enabled individuals to survive after being blown from the mainland [Figure 1, see @Zalucki2004; @Pierce2014; @Hemstrom2022a]. Monarchs are not known to be migratory on any Pacific islands, and no migratory behavior was observed in Australia until the 1930s, when non-reproductive overwintering aggregations were observed in northern New South Wales [NSW, see @James2019]. These and other individuals were later shown to migrate hundreds of kilometers each year to their wintering grounds from more southerly, inland locations in NSW [@James2019;@jamesInductionReproductiveDormancy1983]. Thus, more than 90 years (and hundreds of generations) after expanding out of North America, and more than 90 years without any evidence for seasonal migration, migration re-emerged in these populations. Understanding how this happened could provide crucial context around the "extinction" of migration in monarchs and other species.

To better understand how migration re-emerged in Australian monarchs, we studied continuously breeding monarch populations in Queensland. While these populations do not experience substantial seasonal fluctuations in resource availability, and thus do not migrate [@James2019], recent work has shown that reproductive diapause can be induced in some, *but not all* individuals from these populations if they are exposed to a reduction in photoperiod during development [@Freedman2018], a trait that these non-migratory populations do not display in the wild. In brief, reproductive diapause, where individuals delay their reproductive maturation, is a critical and fitness-relevant phenotype for migratory monarchs because it allows them to invest more heavily in lipid storage and increase their average longevity [@Herman2001;@denlingerInsectDiapauseRich2023;@numataDiapauseUnivoltineSemivoltine2023]. Like other migratory behaviors, diapause is initiated by changes in day length, temperature, and other seasonal fluctuations [@Goehring2002;@iiamsPhotoperiodicClockRegulation2019;@saundersDormancyDiapauseRole2020] and is thought to be directly controlled by juvenile hormone (JH) titers during development [@GreenII2019;@Herman2001;@denlingerInsectDiapauseRich2023].

While diapause is one of many migration-assoicated traits (such as wing morphology, resting and flight metabolic rates, lifespan, and lipid storage), it in particular is linked extremely tightly to migration. Nearly all wild migratory monarch females are in diapause: previous work has found that between 82 and 100% of female migratory monarch butterflies are non-reproductive [@majewskaExposureNonNativeTropical2019; @satterfieldMigratoryMonarchsThat2018a; @Goehring2002; @hermanStudiesAdultReproductive1981]. In contrast, previous rearing trials under typical North Americam summer breeding conditions in have yielded nearly 100% reproductively mature females [@Goehring2002]. These trends are less pronounced in males monarchs, which are therefore not the focus of this study [@majewskaExposureNonNativeTropical2019; @satterfieldMigratoryMonarchsThat2018a]. Given this, diapause serves as an excellent proxy for migration and has been used as such or as a focal trait for the study of migration in monarchs in many studies [see @satterfieldMigratoryMonarchsThat2018; @Goehring2002; @Tenger-Trolander2019; @GreenII2019; @majewskaExposureNonNativeTropical2019; @Freedman2018, for example].

The variance in diapause induction in Queensland monarchs allows for a direct search for genetic associations with this trait and thus for an examination of the genetic underpinnings of the re-evolution of migration in Australian monarchs. Here, we used reduced representation sequencing to search for genes associated with migratory diapause in Queensland monarch butterflies. We found three strongly associated loci, one of which co-located with spectrin beta chain protein involved in actin filament binding, a homolog of which is known to be expressed in silkworm ovaries. This gene had not previously been identified as associated with migration in North American monarch butterflies, and we did not find that any previously identified migratory genes were associated with diapause in our study.

# Materials and Methods

## Background

Although the ancestral range of monarch butterflies is believed to be in North and Central America, they have fairly recently expanded into South America and the Caribbean, and very recently to several locations throughout the Atlantic and Pacific [@oplerFieldGuideWestern1999; @Zhan2014; @haegerMonarchsAtlanticOcean2015]. In North America, the species is well known for its unique, multi-generational migratory life history wherein individuals dispersing northward in the spring do so over three to four generations and then return to their wintering grounds in a single step [@malcolmSpringRecolonizationEastern1993; @cockrellTimeTemperatureLatitudinal1993].

In Southern Florida, the Caribbean, Central America, South America, and throughout most of their introduced range monarchs are non-migratory [@dockxDirectionalStabilizingSelection2007;@altizerPopulationsMonarchButterflies2010;@northMonarchButterflyBiology2004]. Residency in these populations appears to be a derived trait which has arisen multiple times after the species' split from *Danaus erippus*, their closest extant relative [@Zhan2014]. The Pacific expansion of monarchs is relatively recent, with historical records indicating that they established first in Hawaii in the 1840s, likely as a result of individuals blown in during storm events that were able to survive on recently introduced milkweeds [@clarkeMonarchsAustraliaWinds2004; @Pierce2014]. Genetic evidence is consistent with an introduction to Hawaii during this time period [@Hemstrom2022a]. As described above, they then spread across the Pacific and reached Australia in approximately 1871 [@Zalucki2004], and were recorded migrating again in Australia sometime in the 1930s [see Figure 1, @James2019].

## Sample collection, incubation, and phenotyping

@@

## Statistical analysis
We first confirmed the parentage of our samples using the Colony2 program, assuming polygamous, random mating [@Jones2010]. In order to determine the degree to which diapause status was biased between families, we fit a basic linear model in R version 4.2.2 [@rcoreteamLanguageEnvironmentStatistical2022] using phenotypic status as the response variable and maternal family as a fixed effect. Phenotypic status was coded as 0, 1, or 2 for reproductively immature, partially mature, or mature, respectively. We used an ANOVA test in R to determine if maternal ID significantly improved model fit. To determine the expected phenotypic distribution for each family sorted by mean reproductive rank under a null distribution, we randomly permuted individuals between families 10,000 times and then ranked each family for reproductive score.

We conducted a Genome-Wide Association Study (GWAS) in order to determine if any loci were significantly associated with reproductive status. Since our data was generally of low coverage, we used a frequency test in ANGSD [@Korneliussen2014] based on genotype likelihoods rather than called genotypes and thus increase the effective number of loci we were able to analyze. We used the following parameters: -doMajorMinor 1, -SNP_pval 1e-12, -GL 1 (calculate genotype likelihoods using the SAMtools approach), -minQ 20, -minMapQ 20, -minInd 102 (50% of the individuals with called phenotypes), -minMaf 0.05, -doMaf 2 (assume a known minor allele), -doAsso 1 (do an association test using a frequency test), and -yquant coding phenotypic status as above. Since our samples were composed of many groups of full and half-siblings and came from two genetically different years, we also used the -cov argument and supplied the first 20 principal components derived from PCAngsd [@meisnerInferringPopulationStructure2018], which we ran using the default parameters using genotype likelihoods derived from ANGSD using the same parameters plus -doGlf 2 to save genotype likelihoods. We calculated p-values for the resulting likelihood ratio test score and corrected those values using the false discovery rate approach of Benjamini and Hochberg [-@benjaminiControllingFalseDiscovery1995a] in R.
